Partial Least Squares Regression with Multiple Functional and Scalar Predictors

### 2025-07-10 manuscript/main.tex
gram matrix notation

	\subsubsection{Finite-basis approximation of the predictors and regression coefficients}\label{sec: finite basis approximation}	
	Given an orthonormal basis $\{b_l(t)\}$ of $L^2([0,1])$, the functional predictors and the corresponding regression coefficients can be decomposed into
	\begin{equation*}
		X_{ij}(t) = \sum_{l=1}^\infty \theta_{ijl} \, b_l(t), \quad 
		\beta_j(t) = \sum_{l=1}^\infty \eta_{0,jl} \, b_l(t),
	\end{equation*}
	where $\theta_{ijl}$ and $\eta_{0,jl}$ are the coefficients of $X_{ij}(t)$ and $\beta_j(t)$ corresponding to the $l$th basis function $b_l(t)$, respectively.
	In practice, functional predictors are observed only at a finite set of time points rather than as complete curves. Rather than reconstructing infinite-dimensional object stated above via finite observations, we approximate each predictor using a finite-dimensional projection by truncating its basis expansion at a finite number $M$. This $M$ can be set to a sufficiently large value (e.g., 15 or 20) to capture a wide range of functional patterns. 
	Notably, it is unnecessary to fine-tune $M$, as a  penalization scheme introduced in Section \ref{sec: Regularization of PLS Components} regularizes each functional component to achieve the desired smoothness. We denote the projected predictors and coefficients as
	\begin{equation*}
		\widetilde{X}_{ij}(t) = \sum_{l=1}^M \theta_{ijl} \, b_l(t), \quad 
		\tilde{\beta}_j(t) = \sum_{l=1}^M \eta_{0,jl} \, b_l(t).
	\end{equation*}
	The $i$th  hybrid predictor with projected functional parts, is a tuple, denoted as
	\begin{equation*}
		\widetilde{W}_i := 
		(
		X_{i1}, \ldots, X_{iK}, \boldsymbol{Z}_i
		).
	\end{equation*}
	We adopt the following notion for the coefficients:
	\begin{equation}
		\boldsymbol{\theta}_{ij} := (\theta_{ij1}, \ldots, \theta_{ijM})^\top \in \mathbb{R}^M,
		\quad
		\boldsymbol{\eta}_{0j} := (\eta_{0,j1}, \ldots, \eta_{0,jM})^\top  \in \mathbb{R}^M,
	\end{equation}
	and
	\begin{equation}
		\Theta_j := (\boldsymbol{\theta}_{1j}, \ldots, \boldsymbol{\theta}_{nj})^\top  \in \mathbb{R}^{n \times M}.
	\end{equation}

@@

	The goal is to find a unit-norm direction $\boldsymbol{\xi} \in \mathcal{H}$, whose functional components lie in $\operatorname{span} \bigl( b_1(t), \ldots, b_M(t) \bigr)$, that maximizes the empirical covariance between the resulting PLS component scores $\langle \widetilde{W}_1, \boldsymbol{\xi} \rangle_{\mathcal{H}}, \ldots, \langle \widetilde{W}_n, \boldsymbol{\xi} \rangle_{\mathcal{H}}$ and the responses $y_1, \ldots, y_n$.
	We denote the solution as
	\begin{equation}
		\hat{\boldsymbol{\xi}} 
		= 
		(\hat{\xi}_1, \ldots, \hat{\xi}_K, \hat{\boldsymbol{\zeta}}),
	\end{equation}
	where $\hat{\boldsymbol{\zeta}} \in \mathbb{R}^p$ is the scalar part, and each functional part $\hat{\xi}_j \in L^2([0,1])$ lies in $\operatorname{span}(b_1(t), \ldots, b_M(t))$ and is expressed as
	\begin{equation*}
		\hat{\xi}_j(t) = 
		\sum_{l=1}^M d_{jl} \, b_l(t), \quad j = 1, \ldots, K.
	\end{equation*}

	Let $\mathbf{d}_j := (d_{j1}, \ldots, d_{jM})^\top \in \mathbb{R}^M$ and $\mathbf{y} := (y_1, \ldots, y_n)^\top$. We compute $\mathbf{d}_1, \ldots, \mathbf{d}_K$ and $\hat{\boldsymbol{\zeta}}$ as follows.
	First, we obtain the unnormalized coefficient vectors. For each $j$, the unnormalized vector $\bar{\mathbf{d}}_j$ is computed by solving
	\begin{equation*}
		B \bar{\mathbf{d}}_j = (\Theta_j B)^\top \mathbf{y},
	\end{equation*}
	The unnormalized scalar component is given by
	\begin{equation}
		\bar{\boldsymbol{\zeta}} := Z^\top \mathbf{y}.
	\end{equation}
	Next, we normalize all components with respect to the hybrid Hilbert norm (see Definition~\ref{def:hilbert_space}) to ensure $\hat{\boldsymbol{\xi}}$ has unit length. Specifically, we define
	\begin{equation}\label{def:computed_PLS_direction}
		\mathbf{d}_j := 
		\frac{1}{
			\left(
			\sum_{j=1}^K \bar{\mathbf{d}}_j^\top B \bar{\mathbf{d}}_j +
			\bar{\boldsymbol{\zeta}}^\top \bar{\boldsymbol{\zeta}}
			\right)^{1/2}
		} \bar{\mathbf{d}}_j, \quad j = 1, \ldots, K,
	\end{equation}
	and
	\begin{equation}
		\hat{\boldsymbol{\zeta}} :=
		\frac{1}{
			\left(
			\sum_{j=1}^K \bar{\mathbf{d}}_j^\top B \bar{\mathbf{d}}_j +
			\bar{\boldsymbol{\zeta}}^\top \bar{\boldsymbol{\zeta}}
			\right)^{1/2}
		} \bar{\boldsymbol{\zeta}}.
	\end{equation}

>>>>

	\subsubsection{Finite-basis approximation of the predictors and regression coefficients}\label{sec: finite basis approximation}	
	Given a basis $\{b_l(t)\}$ of $L^2([0,1])$, the functional predictors and the corresponding regression coefficients can be decomposed into
	\begin{equation*}
		X_{ij}(t) = \sum_{l=1}^\infty \theta_{ijl} \, b_l(t), \quad 
		\beta_j(t) = \sum_{l=1}^\infty \eta_{0,jl} \, b_l(t),
	\end{equation*}
	where $\theta_{ijl}$ and $\eta_{0,jl}$ are the coefficients of $X_{ij}(t)$ and $\beta_j(t)$ corresponding to the $l$th basis function $b_l(t)$, respectively.
	In practice, functional predictors are observed only at a finite set of time points rather than as complete curves. Rather than reconstructing infinite-dimensional object stated above via finite observations, we approximate each predictor using a finite-dimensional projection by truncating its basis expansion at a finite number $M$. This $M$ can be set to a sufficiently large value (e.g., 15 or 20) to capture a wide range of functional patterns. 
	Notably, it is unnecessary to fine-tune $M$, as a  penalization scheme introduced in Section \ref{sec: Regularization of PLS Components} regularizes each functional component to achieve the desired smoothness. We denote the projected predictors and coefficients as
	\begin{equation*}
		\widetilde{X}_{ij}(t) = \sum_{l=1}^M \theta_{ijl} \, b_l(t), \quad 
		\tilde{\beta}_j(t) = \sum_{l=1}^M \eta_{0,jl} \, b_l(t).
	\end{equation*}
	The $i$th  hybrid predictor with projected functional parts, is a tuple, denoted as
	\begin{equation*}
		\widetilde{W}_i := 
		(
		X_{i1}, \ldots, X_{iK}, \boldsymbol{Z}_i
		).
	\end{equation*}
	We adopt the following notation for the coefficients:
	\begin{equation}
		\boldsymbol{\theta}_{ij} := (\theta_{ij1}, \ldots, \theta_{ijM})^\top \in \mathbb{R}^M,
		\quad
		\boldsymbol{\eta}_{0j} := (\eta_{0,j1}, \ldots, \eta_{0,jM})^\top  \in \mathbb{R}^M,
	\end{equation}
	and
	\begin{equation}
		\Theta_j := (\boldsymbol{\theta}_{1j}, \ldots, \boldsymbol{\theta}_{nj})^\top  \in \mathbb{R}^{n \times M}.
	\end{equation}

@@

	The goal is to find a unit-norm direction $\boldsymbol{\xi} \in \mathcal{H}$, whose functional components lie in $\operatorname{span} \bigl( b_1(t), \ldots, b_M(t) \bigr)$, that maximizes the empirical covariance between the resulting PLS component scores $\langle \widetilde{W}_1, \boldsymbol{\xi} \rangle_{\mathcal{H}}, \ldots, \langle \widetilde{W}_n, \boldsymbol{\xi} \rangle_{\mathcal{H}}$ and the responses $y_1, \ldots, y_n$.
	We denote the solution as
	\begin{equation}
		\hat{\boldsymbol{\xi}} 
		= 
		(\hat{\xi}_1, \ldots, \hat{\xi}_K, \hat{\boldsymbol{\zeta}}),
	\end{equation}
	where $\hat{\boldsymbol{\zeta}} \in \mathbb{R}^p$ is the scalar part, and each functional part $\hat{\xi}_j \in L^2([0,1])$ lies in $\operatorname{span}(b_1(t), \ldots, b_M(t))$ and is expressed as
	\begin{equation*}
		\hat{\xi}_j(t) := 
		\sum_{l=1}^M d_{jl} \, b_l(t), \quad j = 1, \ldots, K.
	\end{equation*}

	Let $\mathbf{d}_j := (d_{j1}, \ldots, d_{jM})^\top \in \mathbb{R}^M$ and $\mathbf{y} := (y_1, \ldots, y_n)^\top$. 
	Let $B \in \mathbb{R}^{M \times M}$ denote the Gram matrix of the basis functions, with entries
	\[
	B_{l, l'} := \langle b_l, b_{l'} \rangle_{L^2([0,1])}, \quad l, l' = 1, \ldots, M.
	\]
	We compute $\mathbf{d}_1, \ldots, \mathbf{d}_K$ and $\hat{\boldsymbol{\zeta}}$ as follows.
	First, we obtain the unnormalized coefficient vectors. For each $j$, the unnormalized vector $\bar{\mathbf{d}}_j$ is computed by solving
	\begin{equation*}
		B \bar{\mathbf{d}}_j = (\Theta_j B)^\top \mathbf{y},
	\end{equation*}
	The unnormalized scalar component is given by
	\begin{equation}
		\bar{\boldsymbol{\zeta}} := Z^\top \mathbf{y}.
	\end{equation}
	Next, we normalize all components with respect to the hybrid Hilbert norm (see Definition~\ref{def:hilbert_space}) to ensure $\hat{\boldsymbol{\xi}}$ has unit length. Specifically, we define
	\begin{equation}\label{def:computed_PLS_direction}
		\mathbf{d}_j := 
		\frac{1}{
			\left(
			\sum_{j=1}^K \bar{\mathbf{d}}_j^\top B \bar{\mathbf{d}}_j +
			\bar{\boldsymbol{\zeta}}^\top \bar{\boldsymbol{\zeta}}
			\right)^{1/2}
		} \bar{\mathbf{d}}_j, \quad j = 1, \ldots, K,
	\end{equation}
	and
	\begin{equation}
		\hat{\boldsymbol{\zeta}} :=
		\frac{1}{
			\left(
			\sum_{j=1}^K \bar{\mathbf{d}}_j^\top B \bar{\mathbf{d}}_j +
			\bar{\boldsymbol{\zeta}}^\top \bar{\boldsymbol{\zeta}}
			\right)^{1/2}
		} \bar{\boldsymbol{\zeta}}.
	\end{equation}
